---
title: "Prediction Project"
author: "HaiyanChen"
date: "2026-04-21"
output:
  pdf_document:
    latex_engine: xelatex
---

## Overview
This report focuses on developing a predictive model based on accelerometer data collected from sensors placed on the belt, forearm, arm, and dumbbell. Model performance is evaluated using cross-validation and appropriate error metrics. The final model is then applied to predict outcomes for 20 unseen test cases.

## Download data
```{r download_data, echo = TRUE}
setwd("E:/coursera_temporary/quiz_environment")

# Download training file
fileUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
download.file(fileUrl,destfile="E:/coursera_temporary/Practical_Machine_Learning/training.csv",method="curl")

# Download testing file
fileUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(fileUrl,destfile="E:/coursera_temporary/Practical_Machine_Learning/testing.csv",method="curl")

# Read data into R
training = read.csv("E:/coursera_temporary/Practical_Machine_Learning/training.csv")
testing = read.csv("E:/coursera_temporary/Practical_Machine_Learning/testing.csv")
```

## Data cleaning
```{r data_cleaning, echo = TRUE, results = "hide", warning = FALSE}
# Load packages
library(caret)
library(caretEnsemble)

# 1. Convert empty strings to NA
training[training == ""] <- NA
testing[testing == ""] <- NA

# 2. Remove high-NA variables
na_ratio <- sapply(training, function(x) mean(is.na(x)))
training_clean <- training[, na_ratio < 0.95]

# 3. Remove irrelevant columns
training_clean <- training_clean[, !(names(training_clean) %in% c(
  "X", "user_name",
  "raw_timestamp_part_1", "raw_timestamp_part_2",
  "cvtd_timestamp", "new_window", "num_window"
))]

# 4. Define predictors (IMPORTANT FIX)
predictor_names <- setdiff(names(training_clean), "classe")

# keep only variables that also exist in testing (SAFE FIX)
predictor_names <- predictor_names[predictor_names %in% names(testing)]

# 5. Align testing dataset (FIXED)
testing_clean <- testing[, predictor_names, drop = FALSE]
```

## Train-test split (for model evaluation ONLY)
```{r train_test_split, echo = TRUE}
set.seed(125)
inTrain <- createDataPartition(training_clean$classe, p = 0.7, list = FALSE)

train_split <- training_clean[inTrain, ]
test_split  <- training_clean[-inTrain, ]
```

## Cross-validation setup
```{r cross_validation_setup, echo = TRUE, results = "hide"}
set.seed(125)  # Set seed for reproducibility

control <- trainControl(
  method = "cv",                 # Use cross-validation
  number = 5,                    # Number of folds (5-fold CV)
  savePredictions = "final",     # Save CV predictions (required for stacking)
  classProbs = TRUE              # Calculate class probabilities
)
```

## Train models
```{r train_model, echo = TRUE, results = "hide"}
model_list <- caretList(
  classe ~ .,                  # Predict Class using all predictors
  data =  train_split,            # Training dataset
  trControl = control,        # Cross-validation settings
  methodList = c("rpart", "rf", "gbm")  # Models: decision tree, random forest, GBM
)
```

## Stacking model
```{r stacking_model, echo = TRUE, results = "hide"}
stack_model <- caretStack(
  model_list,                                   # Base models to combine
  method = "rf",                                # Use random forest as meta-model
  trControl = trainControl(method = "cv", number = 5)  # Cross-validation for stacking
)
```

## Model evaluation (VALIDATION SET ONLY)
In this part, we will evaluate the pre-built models (rpart, rf, gbm) on the validation set (test_split) , which was held out from the training data. Based on their accuracy scores, we will identify the best-performing model.
```{r model_evaluation, echo = TRUE}
# check the class of test_split$classe 
# Ensure it is a factor variable so that it has levels
class(test_split$classe)
 
# Convert test_split$classe to factor 
# Convert test_split$classe to factor 
test_split$classe <- factor(test_split$classe)

# Get base model predictions
pred_rpart <- predict(model_list$rpart, test_split)
pred_rf <- predict(model_list$rf, test_split)
pred_gbm <- predict(model_list$gbm, test_split)

# Get stacking model predictions
pred_stack_raw <- predict(stack_model, test_split)

# Debug: Check what we got
cat("=== Debug Information ===\n")
cat("Class of pred_stack_raw:", class(pred_stack_raw), "\n")

if (is.matrix(pred_stack_raw)) {
  cat("Dimensions:", dim(pred_stack_raw), "\n")
  cat("Column names:", colnames(pred_stack_raw), "\n")
}

# Fix: Convert to single vector of class labels
if (is.matrix(pred_stack_raw)) {
  # If it's a probability matrix (samples × classes)
  if (ncol(pred_stack_raw) == 5) {
    # Convert probabilities to class labels
    pred_stack <- colnames(pred_stack_raw)[max.col(pred_stack_raw)]
    cat("Converted probability matrix to classes\n")
  } else {
    # If it's already a matrix but not probabilities
    pred_stack <- as.vector(pred_stack_raw[, 1])
  }
} else if (is.data.frame(pred_stack_raw)) {
  # If it's a data frame
  if (ncol(pred_stack_raw) == 5) {
    pred_stack <- colnames(pred_stack_raw)[max.col(pred_stack_raw)]
  } else {
    pred_stack <- pred_stack_raw[, 1]
  }
} else {
  # If it's already a vector
  pred_stack <- pred_stack_raw
}

# Verify lengths after conversion
cat("\nLength after conversion:\n")
cat("test_split$classe:", length(test_split$classe), "\n")
cat("pred_stack:", length(pred_stack), "\n")

# Force same factor levels
target_levels <- levels(test_split$classe)
pred_rpart <- factor(pred_rpart, levels = target_levels)
pred_rf <- factor(pred_rf, levels = target_levels)
pred_gbm <- factor(pred_gbm, levels = target_levels)
pred_stack <- factor(pred_stack, levels = target_levels)

# Now create confusion matrices
cm_rpart <- confusionMatrix(pred_rpart, test_split$classe)
cm_rf <- confusionMatrix(pred_rf, test_split$classe)
cm_gbm <- confusionMatrix(pred_gbm, test_split$classe)
cm_stack <- confusionMatrix(pred_stack, test_split$classe)

# Combine results
accuracy_compare <- data.frame(
  Model = c("rpart", "rf", "gbm", "stacking"),
  Accuracy = c(
    cm_rpart$overall["Accuracy"],
    cm_rf$overall["Accuracy"],
    cm_gbm$overall["Accuracy"],
    cm_stack$overall["Accuracy"]
  ),
  CI_Lower = c(
    cm_rpart$overall["AccuracyLower"],
    cm_rf$overall["AccuracyLower"],
    cm_gbm$overall["AccuracyLower"],
    cm_stack$overall["AccuracyLower"]
  ),
  CI_Upper = c(
    cm_rpart$overall["AccuracyUpper"],
    cm_rf$overall["AccuracyUpper"],
    cm_gbm$overall["AccuracyUpper"],
    cm_stack$overall["AccuracyUpper"]
  ),
  P_Value = c(
    cm_rpart$overall["AccuracyPValue"],
    cm_rf$overall["AccuracyPValue"],
    cm_gbm$overall["AccuracyPValue"],
    cm_stack$overall["AccuracyPValue"]
  )
)

# Print comparison table
print(accuracy_compare)

# Find and display the best model
best <- accuracy_compare[which.max(accuracy_compare$Accuracy), ]
cat("\n===== Best Model =====\n")
cat("Model:", best$Model, "\n")
cat("Accuracy:", round(best$Accuracy, 4), "\n")
cat("95% CI: [", round(best$CI_Lower, 4), ",", round(best$CI_Upper, 4), "]\n")
cat("P-value:", format(best$P_Value, scientific = TRUE), "\n")
```
According to the result, the best model is `r best$Model`, with the largest accuracy `r round(best$Accuracy, 4)` (95% CI: `r round(best$CI_Lower, 4)` to `r round(best$CI_Upper, 4)`).

## Final prediction
Finally, we use the best model (`r best$Model`) to predict the testing clean dataset.
```{r final_prediction, echo = TRUE, results = "hide"}
final_pred <- predict(stack_model, testing_clean)

results <- data.frame(
  problem_id = testing$problem_id, 
  classe = final_pred
)

results
```
